---
title: "caQTL_cTWAS"
author: "Lifan Liang"
date: "2023-12-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Peak selection

cPeaks was defined as FDR < 0.05. The inclusion of peaks for each condition (a unique combination of time points and cell types) is: (1) a caPeak in any time point in the same cell type; (2) nominal P value < 0.1 in the current condition. The number of peaks was shown in the table below

```{r, echo=F}
GABA1 <- read.table("data/GABA_peakonly_res.txt", header=T)
nmglut1 <- read.table("data/nmglut_peakonly_res.txt", header=T)
npglut1 <- read.table("data/npglut_peakonly_res.txt", header=T)

npeaks <- rbind(table(GABA1$timepoint), table(nmglut1$timepoint), table(npglut1$timepoint))
colnames(npeaks) <- paste0(c(0,1,6),"hour")
rownames(npeaks) <- c("GABA", "nmglut", "npglut")
knitr::kable(npeaks, caption="Number of peaks included in cTWAS")
```

## cTWAS parameter estimation

For each peak selected, the lead caQTL was used to build the prediction model. cTWAS first performed parameter estimation without fine mapping.

### GABA parameters
```{r, echo=F}
GABA.params <- readRDS("data/GABA_params.rds")
nmglut.params <- readRDS("data/nmglut_params.rds")
npglut.params <- readRDS("data/npglut_params.rds")

GABA.params$convergence_plot
```

### NPglut parameters
```{r, echo=F}
npglut.params$convergence_plot
```

### NMglut parameters
```{r, echo=F}
nmglut.params$convergence_plot
```

## Significant peaks

A peak is significant for the cell type if its sum of PIP across time points is more than 0.8.

### GABA significant peaks

```{r, echo=F}
library(kableExtra)
options(knitr.kable.NA = '')
GABA2 <- aggregate(GABA1$susie_pip, list(GABA1$genename), sum)
GABA3 <- merge(GABA2, GABA1[GABA1$timepoint=="0hr_GABA",c("genename","susie_pip")], 
                 by.x="Group.1", by.y="genename",all.x = T)
colnames(GABA3) <- c("genename","PIP_sum","PIP_0hr_GABA")
GABA4 <- merge(GABA3, GABA1[GABA1$timepoint=="1hr_GABA",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(GABA4)[4] <- "PIP_1hr_GABA"
GABA5 <- merge(GABA4, GABA1[GABA1$timepoint=="6hr_GABA",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(GABA5)[5] <- "PIP_6hr_GABA"
GABA5 <- GABA5[order(GABA5$PIP_sum, decreasing = T),]
GABA5$PIP_diff <- GABA5$PIP_sum - 2 * GABA5$PIP_0hr_GABA
colnames(GABA5)[1] <- "PeakRange"
GABA5 <- GABA5[order(GABA5$PIP_sum,decreasing = T),]
GABA5[GABA5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the GABA cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

### npglut significant peaks

```{r, echo=F}
options(knitr.kable.NA = '')
npglut2 <- aggregate(npglut1$susie_pip, list(npglut1$genename), sum)
npglut3 <- merge(npglut2, npglut1[npglut1$timepoint=="0hr_npglut",c("genename","susie_pip")], 
                 by.x="Group.1", by.y="genename",all.x = T)
colnames(npglut3) <- c("genename","PIP_sum","PIP_0hr_npglut")
npglut4 <- merge(npglut3, npglut1[npglut1$timepoint=="1hr_npglut",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(npglut4)[4] <- "PIP_1hr_npglut"
npglut5 <- merge(npglut4, npglut1[npglut1$timepoint=="6hr_npglut",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(npglut5)[5] <- "PIP_6hr_npglut"
npglut5 <- npglut5[order(npglut5$PIP_sum, decreasing = T),]
npglut5$PIP_diff <- npglut5$PIP_sum - 2 * npglut5$PIP_0hr_npglut
npglut5 <- npglut5[order(npglut5$PIP_sum,decreasing = T),]
colnames(npglut5)[1] <- "PeakRange"
npglut5[npglut5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the npglut cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

### nmglut significant peaks

```{r, echo=F}
nmglut2 <- aggregate(nmglut1$susie_pip, list(nmglut1$genename), sum)
nmglut3 <- merge(nmglut2, nmglut1[nmglut1$timepoint=="0hr_nmglut",c("genename","susie_pip")], 
               by.x="Group.1", by.y="genename",all.x = T)
colnames(nmglut3) <- c("genename","PIP_sum","PIP_0hr_nmglut")
nmglut4 <- merge(nmglut3, nmglut1[nmglut1$timepoint=="1hr_nmglut",c("genename","susie_pip")], 
               by="genename",all.x = T)
colnames(nmglut4)[4] <- "PIP_1hr_nmglut"
nmglut5 <- merge(nmglut4, nmglut1[nmglut1$timepoint=="6hr_nmglut",c("genename","susie_pip")], 
               by="genename",all.x = T)
colnames(nmglut5)[5] <- "PIP_6hr_nmglut"
nmglut5$PIP_diff <- nmglut5$PIP_sum - 2 * nmglut5$PIP_0hr_nmglut
nmglut5 <- nmglut5[order(nmglut5$PIP_sum,decreasing = T),]
colnames(nmglut5)[1] <- "PeakRange"
nmglut5[nmglut5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the nmglut cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

## Cell type overlap

```{r, echo=F, message=F}
library(ggvenn)

ggvenn(list(GABA=GABA5$PeakRange, nmglut=nmglut5$PeakRange, npglut=npglut5$PeakRange),
       c("GABA","nmglut","npglut"))
```



