---
title: "caQTL_cTWAS"
author: "Lifan Liang"
date: "2023-12-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Peak selection

cPeaks was defined as FDR < 0.05. The inclusion of peaks for each condition (a unique combination of time points and cell types) is: (1) a caPeak in any time point in the same cell type; (2) nominal P value < 0.1 in the current condition. The number of peaks was shown in the table below

```{r, echo=F}
GABA1 <- readRDS("data/caQTL_GABA_res.RDS")
nmglut1 <- readRDS("data/caQTL_nmglut_res.RDS")
npglut1 <- readRDS("data/caQTL_npglut_res.RDS")

npeaks <- rbind(table(GABA1$timepoint), table(nmglut1$timepoint), table(npglut1$timepoint))
colnames(npeaks) <- paste0(c(0,1,6),"hour")
rownames(npeaks) <- c("GABA", "nmglut", "npglut")
knitr::kable(npeaks, caption="Number of peaks included in cTWAS")
```

## cTWAS parameter estimation

For each peak selected, the lead caQTL was used to build the prediction model. cTWAS first performed parameter estimation without fine mapping.

### GABA parameters
```{r, echo=F}
GABA.params <- readRDS("data/GABA_params.rds")
nmglut.params <- readRDS("data/nmglut_params.rds")
npglut.params <- readRDS("data/npglut_params.rds")

GABA.params$convergence_plot
```

```{r}
library(RColorBrewer)
myPalette <- brewer.pal(4, "Set2") 
pie(GABA.params$group_pve, col=myPalette, border="white",
    main = paste0(sprintf("%.2f", 100-(GABA.params$group_pve[1]/GABA.params$total_pve)*100), "% PVE for GABA peaks"))
```

### NPglut parameters
```{r, echo=F}
npglut.params$convergence_plot
```

```{r, echo=F}
pie(npglut.params$group_pve, col=myPalette, border="white",
    main = paste0(sprintf("%.2f", 100-(npglut.params$group_pve[1]/npglut.params$total_pve)*100), "% PVE for NPglut peaks"))
```

### NMglut parameters

```{r, echo=F}
nmglut.params$convergence_plot
```

```{r, echo=F}
pie(nmglut.params$group_pve, col=myPalette, border="white",
    main = paste0(sprintf("%.2f", 100-(nmglut.params$group_pve[1]/nmglut.params$total_pve)*100), "% PVE for NMglut peaks"))
```


## Significant peaks

A peak is significant for the cell type if its sum of PIP across time points is more than 0.8. There are 22 peaks in GABA, 6 peaks in NPglut, and 11 peaks in NMglut.

### GABA significant peaks

```{r, echo=F}
library(kableExtra)
options(knitr.kable.NA = '')
GABA2 <- aggregate(GABA1$susie_pip, list(GABA1$genename), sum)
GABA3 <- merge(GABA2, GABA1[GABA1$timepoint=="0hr_GABA",c("genename","susie_pip")], 
                 by.x="Group.1", by.y="genename",all.x = T)
colnames(GABA3) <- c("genename","PIP_sum","PIP_0hr_GABA")
GABA4 <- merge(GABA3, GABA1[GABA1$timepoint=="1hr_GABA",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(GABA4)[4] <- "PIP_1hr_GABA"
GABA5 <- merge(GABA4, GABA1[GABA1$timepoint=="6hr_GABA",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(GABA5)[5] <- "PIP_6hr_GABA"
GABA5[is.na(GABA5)] <- 0
GABA5 <- GABA5[order(GABA5$PIP_sum, decreasing = T),]
GABA5$PIP_diff <- GABA5$PIP_sum - 2 * GABA5$PIP_0hr_GABA
colnames(GABA5)[1] <- "PeakRange"
GABA5 <- GABA5[order(GABA5$PIP_sum,decreasing = T),]
GABA5[GABA5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the GABA cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

### npglut significant peaks

```{r, echo=F}
options(knitr.kable.NA = '')
npglut2 <- aggregate(npglut1$susie_pip, list(npglut1$genename), sum)
npglut3 <- merge(npglut2, npglut1[npglut1$timepoint=="0hr_npglut",c("genename","susie_pip")], 
                 by.x="Group.1", by.y="genename",all.x = T)
colnames(npglut3) <- c("genename","PIP_sum","PIP_0hr_npglut")
npglut4 <- merge(npglut3, npglut1[npglut1$timepoint=="1hr_npglut",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(npglut4)[4] <- "PIP_1hr_npglut"
npglut5 <- merge(npglut4, npglut1[npglut1$timepoint=="6hr_npglut",c("genename","susie_pip")], 
                 by="genename",all.x = T)
colnames(npglut5)[5] <- "PIP_6hr_npglut"
npglut5[is.na(npglut5)] <- 0
npglut5 <- npglut5[order(npglut5$PIP_sum, decreasing = T),]
npglut5$PIP_diff <- npglut5$PIP_sum - 2 * npglut5$PIP_0hr_npglut
npglut5 <- npglut5[order(npglut5$PIP_sum,decreasing = T),]
colnames(npglut5)[1] <- "PeakRange"
npglut5[npglut5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the npglut cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

### nmglut significant peaks

```{r, echo=F}
nmglut2 <- aggregate(nmglut1$susie_pip, list(nmglut1$genename), sum)
nmglut3 <- merge(nmglut2, nmglut1[nmglut1$timepoint=="0hr_nmglut",c("genename","susie_pip")], 
               by.x="Group.1", by.y="genename",all.x = T)
colnames(nmglut3) <- c("genename","PIP_sum","PIP_0hr_nmglut")
nmglut4 <- merge(nmglut3, nmglut1[nmglut1$timepoint=="1hr_nmglut",c("genename","susie_pip")], 
               by="genename",all.x = T)
colnames(nmglut4)[4] <- "PIP_1hr_nmglut"
nmglut5 <- merge(nmglut4, nmglut1[nmglut1$timepoint=="6hr_nmglut",c("genename","susie_pip")], 
               by="genename",all.x = T)
colnames(nmglut5)[5] <- "PIP_6hr_nmglut"
nmglut5[is.na(nmglut5)] <- 0
nmglut5$PIP_diff <- nmglut5$PIP_sum - 2 * nmglut5$PIP_0hr_nmglut
nmglut5 <- nmglut5[order(nmglut5$PIP_sum,decreasing = T),]
colnames(nmglut5)[1] <- "PeakRange"
nmglut5[nmglut5$PIP_sum>0.8,] %>%
  kable(caption = "Significant peaks for the nmglut cell type", digits=3, row.names = F) %>%
  scroll_box(height="400px")
```

## Cell type overlap

### Overlap of the cTWAS input

```{r, echo=F, message=F}
library(ggvenn)

ggvenn(list(GABA=GABA5$PeakRange, nmglut=nmglut5$PeakRange, npglut=npglut5$PeakRange),
       c("GABA","nmglut","npglut"))
```

### Overlap of the cTWAS significant peaks

```{r, echo=F, message=F}
library(ggvenn)

ggvenn(list(GABA=GABA5[GABA5$PIP_sum>0.8,"PeakRange"], nmglut=nmglut5[nmglut5$PIP_sum>0.8,"PeakRange"], npglut=npglut5[npglut5$PIP_sum>0.8,"PeakRange"]), c("GABA","nmglut","npglut"))
```

## Dynamic cPeaks

In addition to PIP sum > 0.8, dynamic cpeaks must satisfy that:

$$
PIP_{GABA6} + PIP_{GABA1} - PIP_{GABA0} \geq 0.5
$$
for the GABA cell type. The same condition applies to the other cell types .

Peaks not included in one time point are regarded as $PIP=0$.

```{r,echo=F,message=F}
library(ggplot2)
Negenes <- c(sum((GABA5$PIP_diff>=0.5) & (GABA5$PIP_sum>0.8)),
             sum((nmglut5$PIP_diff>=0.5) & (nmglut5$PIP_sum>0.8)),
             sum((npglut5$PIP_diff>=0.5) & (npglut5$PIP_sum>0.8)),
             sum((GABA5$PIP_diff<0.5) & (GABA5$PIP_sum>0.8)),
             sum((nmglut5$PIP_diff<0.5) & (nmglut5$PIP_sum>0.8)),
             sum((npglut5$PIP_diff<0.5) & (npglut5$PIP_sum>0.8)))
dat <- data.frame(Ngenes=Negenes, celltype=rep(c("GABA","nmglut","npglut"),2),
           dynamic=c(rep(T,3), rep(F,3)))
ggplot(dat, aes(x=celltype, fill=dynamic, y=Ngenes, label=Ngenes)) + 
  geom_bar(position="stack", stat="identity") +
  geom_text(size = 6, position = position_stack(vjust = 0.5)) + 
  xlab("Number of Top genes") + ylab("Cell type")
```

## Cell type correlation

#### Comparing top genes across cell types nominated in the same locus

This table shows how different cell types may nominate different genes in the same locus. 

All shared peaks are consistent except locus 3&4 in chr7.

```{r, echo=F}
tpeak <- read.table("data/multigroup_cTWAS_toppeak_by_locus.txt", header=T, sep=" ")
tpeak$locus_tag <- rownames(tpeak)
rownames(tpeak) <- NULL
tpeak[,c(4,1,2,3)] %>%
  mutate_if(is.numeric, format, digits=3,nsmall = 0) %>%
  kable(caption="Dynamic genes by locus") %>%
  scroll_box(height="400px")
#knitr::kable(npglut.pip1, caption = "Top genes for the npGlut cell type")
```